Inferring Galactic Parameters from Chemical Abundances with Simulation-Based Inference

### 2025-02-19 src/tex/ms.tex
[showyourwork] overleaf sync

For example, a range of high-mass IMF slopes have been suggested \citep[Tab.\,7]{2016ApJ...824...82C}, with a steeper-than-canonical slope being suggested by a range of studies \citep[e.g.][]{2015ApJ...806..198W,Rybizki2015,Chabrier2014}. In addition, the IMF slope may itself be not a constant but rather a function of metallicity, introducing further complexity \citep[e.g.][]{2019MNRAS.482..118G,Martin2019}. Similarly, the choice of SN\,Ia delay-time-distribution and normalization plays a crucial role in the enrichment of the ISM \citep[e.g.][]{Buck2021} and is heavily debated \citep{2010ApJ...722.1879M,2012MNRAS.426.3282M,2015ApJ...810..137J}.

@@

\paragraph{\texttt{CHEMPY} parameters}
In this work, we allow six \texttt{CHEMPY} parameters to vary freely (see also Tab.\,\ref{tab:priors}). These can be categorized into three groups:
\begin{enumerate}
     \item $\vec\Lambda$: \textbf{Global Galactic Parameters} describe SSP physics and comprise the high-mass \citet{2003PASP..115..763C} IMF slope, $\alpha_\mathrm{IMF}$, and (logarithmic) Type Ia SN normalization, $\log_{10}(N_\mathrm{Ia})$. We treat these as star-independent and assume them to be constant across galactic environments and cosmic time\footnote{Whilst $\log_{10}(N_\mathrm{Ia})$ is constant with respect to time by definition, it being simply a normalization constant, there is some evidence for $\alpha_\mathrm{IMF}$ varying as a function of time or metallicity \citep{Chabrier2014,2016MNRAS.462.2832C,2019MNRAS.482..118G,Martin2019}.}. 
     We adopt the same broad priors as \citep{Philcox_2019} for these variables (see also Tab.\ref{tab:priors}). 
     %
     \item $\{\vec\Theta_i\}$: \textbf{Local Galactic Parameters} describe the local physics of the ISM and are hence specific to each stellar environment, indexed by $i$. As defined in \citep{Rybizki_2017}, these include the star-formation efficiency (SFE), $\log_{10}(\text{SFE})$, the peak of the star formation rate (SFR), $\log_{10}(\mathrm{SFR}_\mathrm{peak})$, and the fraction of stellar outflow that is fed to the gas reservoir, $\mathrm{x}_\mathrm{out}$. We adopt broad priors for all parameters and, as in \citep{Philcox_2019}, fix the SN\,Ia delay-time distribution, $\log_{10}(\tau_{\rm Ia})$, to $\log_{10}(\tau_{\rm Ia})=-0.80$ \citep[see also][]{Philcox_2018}.
     %
     \item $\{T_i\}$: \textbf{Stellar Birth-Times}. Time in Gyr at which a given star is formed from the ISM. We assume that its proto-stellar abundances match the local ISM abundances at $T_i$.
\end{enumerate}

Following \citet{Philcox_2019}, to avoid unrealistic star formation histories (that are very `bursty' for early stars), we additionally require that the SFR (parametrized by a $\Gamma$ distribution with shape parameter $a=2$) at the maximum possible stellar birth-time ($13.8$\,Gyr) should be at least 5\% of the mean SFR, ensuring that there is still a reasonable chance of forming a star at this time-step. This corresponds to the constraint $\log_{10}\left(\mathrm{SFR}_\mathrm{peak}\right)>0.294$. For this reason, a truncated Normal prior will be used for the SFR parameter. Furthermore, we constrain $T_i$ to the interval $[1,13.8]$\,Gyr (assuming an age of the Universe of 13.8\,Gyr), ignoring any stars formed before $1$\,Gyr, which is justified as these are expected to be rare.

@@

\paragraph{Nucleosynthetic yield tables}
We adopt the same nucleosynthetic yield tables as in \citep{Philcox_2019}, see their Sec.~2.2 for more details.
To test our method, we aim further at inferring parameters from a sample of stars taken from a hydrodynamical simulation of a MW type galaxy which we take from the IllustrisTNG project \citep{Pillepich2018}. To ensure maximal compatibility with TNG, we adopt their nucleosynthetic yield tables in \texttt{CHEMPY}, for enrichment by SN\,Ia, SN\,II and AGB stars. The utilized yields are summarized in Tab.\,\ref{tab:chempy_TNG_yields}, matching \citet[Tab.\,2]{2018MNRAS.473.4077P}, and we note that the SN\,II yields are renormalized such that the IMF-weighted yield ratios at each metallicity are equal to those from the \citet{2006ApJ...653.1145K} mass range models alone. \texttt{CHEMPY} uses only net yields, such that they provide only newly synthesized material, with the remainder coming from the initial SSP composition. These tables may not well-represent true stellar chemistry, and the effects of this mismatch are examined in Sec.\,\ref{subsec:mocks_wrong_yield} by performing inference using an alternative set of yields that does not match the yield set of the training data. For the analysis of observational data, we would want to use the most up-to-date yields, such as \citet{2016ApJ...825...26K} AGB yields, and carefully choose elements which are known to be well reproduced by our current models (e.g. shown by \citet{2019ApJ...874..102W,2019arXiv190806113G}), though this is not appropriate in our context. To facilitate best comparison with Ilustris TNG, we further set the maximum SN\,II mass as $100\,\mathrm{M}_\odot$ (matching the IMF upper mass limit), adopt stellar lifetimes from \citet{portinari} and do not allow for any `hypernovae' \citep[in contrary to][]{2018ApJ...861...40P}).

@@

\paragraph{Chemical elements}
In our analysis we only track nine elements: C, Fe, H, He, Mg, N, Ne, O and Si since these are the only elements traced by TNG. We principally compare the logarithmic abundances [X/Fe] and [Fe/H] defined by 
\begin{equation}
     [\mathrm{X}/\mathrm{Y}] = \log_{10}(N_\mathrm X/N_\mathrm Y)_\mathrm{star} - \log_{10}(N_\mathrm X/N_\mathrm Y)_\odot
\end{equation}
for number fraction $N_\mathrm X$ of element X. Here $\odot$ denotes the solar number fractions of \citet{2009ARA&A..47..481A}. This uses H for normalization, thus we are left with $n_\mathrm{el}=8$ independent elements which must be tracked by \texttt{CHEMPY}\footnote{In observational contexts, it may be more appropriate to compute abundances relative to Mg rather than Fe, as in \citep{2019ApJ...874..102W}, since Mg is only significantly produced by SN\,II and hence a simpler tracer of chemical enrichment.}.

With these modifications, \texttt{CHEMPY} allows for fast prediction of TNG-like chemical abundances for a given set of galactic parameters. It is important to note that the two GCE models have very different parametrizations of galactic physics, with TNG including vastly more effects, thus it is not certain \textit{a priori} how useful \texttt{CHEMPY} will be in emulating the TNG simulation, although its utility was partially demonstrated in \citet{2018ApJ...861...40P}. However,  such a test is necessary to prepare for an inference on real data.

@@

In essence, instead of computing the full model for each input parameter set, we pass the parameters to the NN which predicts the output abundances to high accuracy. As already argued in \cite{Philcox_2019} this has two benefits;
\begin{enumerate}
    \item \textbf{Speed:} The run-time of the \texttt{CHEMPY} function is $\sim1$\,s per input parameter set, which leads to very slow generation of training data for SBI. With the NN emulator, this reduces to $\sim5\times10^{-5}$\,s, and is trivially parallelizable, unlike \texttt{CHEMPY}.
    \item \textbf{Differentiability:} The NN is written in pytorch which allows for automatic differentiation. Additionally the NN has a simple closed-form analytic structure \citep[described in the appendix of][]{Philcox_2019}, unlike the complex \texttt{CHEMPY} model. This allows it to be differentiated, so one can use it to sample via advanced methods such HMC as done in \citep{Philcox_2019}.
\end{enumerate}

Despite the additional complexity introduced by using multiple stellar data-points, our NN simply needs to predict the birth-time abundances for a single star (with index $i$) given a set of six parameters; $\{\vec\Lambda,\vec\Theta_i,T_i\}$. The same NN can be used for all $n_\mathrm{stars}$ stars (and run in parallel), reducing a set of $n_\mathrm{stars}$ runs of \texttt{CHEMPY} to a single matrix computation. With the above network parameter choices, the NN predicts abundances with an absolute percentage error of \variable{output/ape_NN.txt}, which is far below typical observational errors and even smaller away from the extremes of parameter space (see Fig.~\ref{fig:ape_NN}). In fact, we will add additional observational uncertainty to our mock observational data later during training of the neural posteriro estimater network.

>>>>

For example, a range of high-mass IMF slopes have been suggested \citep[Tab.\,7]{2016ApJ...824...82C}, with a steeper-than-canonical slope being suggested by a range of studies \citep[e.g.][]{2015ApJ...806..198W,Rybizki2015,Chabrier2014}. In addition, the IMF slope may itself be not a constant but rather a function of metallicity, introducing further complexity \citep[e.g.][]{2019MNRAS.482..118G,Martin2019}. Similarly, the choice of SN\,Ia delay-time-distribution and normalization plays a crucial role in the enrichment of the interstellar medium \citep[ISM; e.g.][]{Buck2021} and is heavily debated \citep{2010ApJ...722.1879M,2012MNRAS.426.3282M,2015ApJ...810..137J}.

@@

@@

\paragraph{\texttt{CHEMPY} parameters:}
In this work, we allow six \texttt{CHEMPY} parameters to vary freely (see also Tab.\,\ref{tab:priors}). These can be categorized into three groups:
\begin{enumerate}
     \item $\vec\Lambda$: \textbf{Global Galactic Parameters} describe SSP physics and comprise the high-mass \citet{2003PASP..115..763C} IMF slope, $\alpha_\mathrm{IMF}$, which effectively sets the number of CC-SNe a SSP generates and (logarithmic) Type Ia SN normalization, $\log_{10}(N_\mathrm{Ia})$, which controls the total number of SN Is per SSP. We treat these as star-independent and assume them to be constant across galactic environments and cosmic time\footnote{Whilst $\log_{10}(N_\mathrm{Ia})$ is constant with respect to time by definition, it being simply a normalization constant, there is some evidence for $\alpha_\mathrm{IMF}$ varying as a function of time or metallicity \citep{Chabrier2014,2016MNRAS.462.2832C,2019MNRAS.482..118G,Martin2019}.}. 
     We adopt the same broad priors as \citep{Philcox_2019} for these variables (see also Tab.\ref{tab:priors}). 
     %
     \item $\{\vec\Theta_i\}$: \textbf{Local Galactic Parameters} describe the local physics of the ISM and are hence specific to each stellar environment, indexed by $i$. As defined in \citep{Rybizki_2017}, these include the star-formation efficiency (SFE), $\log_{10}(\text{SFE})$, which qunatifies the star formation rate per unit gas, the peak of the star formation rate (SFR), $\log_{10}(\mathrm{SFR}_\mathrm{peak})$, and the fraction of stellar outflow that is fed to the gas reservoir, $\mathrm{x}_\mathrm{out}$. We adopt broad priors for all parameters and, as in \citep{Philcox_2019}, fix the SN\,Ia delay-time distribution, $\log_{10}(\tau_{\rm Ia})$, to $\log_{10}(\tau_{\rm Ia})=-0.80$ \citep[see also][]{Philcox_2018}.
     %
     \item $\{T_i\}$: \textbf{Stellar Birth-Times}. Time in Gyr at which a given star is formed from the ISM. We assume that its proto-stellar abundances match the local ISM abundances at $T_i$.
\end{enumerate}

Following \citet{Philcox_2019}, to avoid unrealistic star formation histories (that are very `bursty' for early stars), we additionally require that the SFR (parametrized by a $\Gamma$ distribution with shape parameter $a=2$\footnote{\texttt{CHEMPY} parametrizes the SFR with a $\Gamma$ distribution
\begin{equation*}
\label{eq:SFR}
\mathrm{SFR}\left(t,k,\vartheta\right)=\frac{1}{\Gamma(k)\vartheta^k}t^{k-1}\exp\left(\frac{-t}{\vartheta}\right), \mathrm{for}\,k=2 \rightarrow \vartheta = \mathrm{SFR}_\mathrm{peak} 
\end{equation*}
where the shape parameter is fixed to $k=2$ such that the scale parameter ($\vartheta$) determines the peak of the SFR}) at the maximum possible stellar birth-time ($13.8$\,Gyr) should be at least 5\% of the mean SFR, ensuring that there is still a reasonable chance of forming a star at this time-step. In our formalism, this corresponds to the constraint $\log_{10}\left(\mathrm{SFR}_\mathrm{peak}\right)>0.294$. For this reason, a truncated Gaussian prior will be used for the SFR parameter. Furthermore, we constrain $T_i$ to the interval $[1,13.8]$\,Gyr (assuming an age of the Universe of 13.8\,Gyr), ignoring any stars formed before $1$\,Gyr, which is justified as these are expected to be rare.

\paragraph{Nucleosynthetic yield tables:}
We adopt the same nucleosynthetic yield tables as in \citep{Philcox_2019}, see their Sec.~2.2 for more details.
To test our method, we aim further at inferring parameters from a sample of stars taken from a hydrodynamical simulation of a MW type galaxy which we take from the IllustrisTNG project \citep{Pillepich2018}. To ensure maximal compatibility with TNG, we adopt their nucleosynthetic yield tables in \texttt{CHEMPY}, for enrichment by SN\,Ia, SN\,II and AGB stars. The utilized yields are summarized in Tab.\,\ref{tab:chempy_TNG_yields}, matching \citet[Tab.\,2]{2018MNRAS.473.4077P}, and we note that the SN\,II yields are renormalized such that the IMF-weighted yield ratios at each metallicity are equal to those from the \citet{2006ApJ...653.1145K} mass range models alone. \texttt{CHEMPY} uses only net yields, such that they provide only newly synthesized material, with the remainder coming from the initial SSP composition. These tables may not well-represent true stellar chemistry, and the effects of this mismatch are examined in Sec.\,\ref{subsec:mocks_wrong_yield} by performing inference using an alternative set of yields that does not match the yield set of the training data. For the analysis of observational data, we would want to use the most up-to-date yields, such as \citet{2016ApJ...825...26K} AGB yields, and carefully choose elements which are known to be well reproduced by our current models (e.g. shown by \citet{2019ApJ...874..102W,2019arXiv190806113G}), though this is not appropriate in our context. To facilitate best comparison with Ilustris TNG, we further set the maximum SN\,II mass as $100\,\mathrm{M}_\odot$ (matching the IMF upper mass limit), adopt stellar lifetimes from \citet{portinari} and do not allow for any `hypernovae' \citep[in contrary to][]{2018ApJ...861...40P}).

@@

\paragraph{Chemical elements:}
In our analysis we only track nine elements: C, Fe, H, He, Mg, N, Ne, O and Si since these are the only elements traced by TNG. We principally compare the logarithmic abundances [X/Fe] and [Fe/H] defined by 
\begin{equation}
     [\mathrm{X}/\mathrm{Y}] = \log_{10}(N_\mathrm X/N_\mathrm Y)_\mathrm{star} - \log_{10}(N_\mathrm X/N_\mathrm Y)_\odot
\end{equation}
for number fraction $N_\mathrm X$ of element X. Here $\odot$ denotes the solar number fractions of \citet{2009ARA&A..47..481A}. As is customary, we use H for normalization, thus we are left with $n_\mathrm{el}=8$ independent elements which must be tracked by \texttt{CHEMPY}\footnote{In observational contexts, it may be more appropriate to compute abundances relative to Mg rather than Fe, as in \citep{2019ApJ...874..102W}, since Mg is only significantly produced by SN\,II and hence a simpler tracer of chemical enrichment.}.

With these modifications, \texttt{CHEMPY} allows to predict TNG-like chemical abundances for a given set of galactic parameters. It is important to note that the two GCE models have very different parametrizations of galactic physics, with TNG including vastly more effects, it is thus not certain \textit{a priori} how useful \texttt{CHEMPY} will be in emulating the TNG simulation, although its utility was partially demonstrated in \citet{2018ApJ...861...40P}. However,  such a test is necessary to prepare for an inference on real data.

@@

In essence, instead of computing the full model for each input parameter set, we pass the parameters to the NN which predicts the output abundances to high accuracy. As already argued in \cite{Philcox_2019} this has two benefits;
\begin{enumerate}
    \item \textbf{Speed:} The run-time of the \texttt{CHEMPY} function is $\sim1$\,s per input parameter set, which leads to very slow generation of training data for SBI. With the NN emulator, this reduces to $\sim5\times10^{-5}$\,s, and is trivially parallelizable, unlike \texttt{CHEMPY}.
    \item \textbf{Differentiability:} The NN is written in pytorch and has additionally a simple closed-form analytic structure \citep[described in the appendix of][]{Philcox_2019}, unlike the complex \texttt{CHEMPY} model. Both aspects allow it to be differentiated either via auto-diff or analytically, so one can use it to sample via advanced methods such HMC as done in \citep{Philcox_2019}.
\end{enumerate}

Despite the additional complexity introduced by using multiple stellar data-points, our NN simply needs to predict the birth-time abundances for a single star (with index $i$) from a given set of six parameters; $\{\vec\Lambda,\vec\Theta_i,T_i\}$. The same NN can be used for all $n_\mathrm{stars}$ stars (and run in parallel), reducing a set of $n_\mathrm{stars}$ runs of \texttt{CHEMPY} to a single matrix computation. With the above network parameter choices, the NN predicts abundances with an absolute percentage error of \variable{output/ape_NN.txt} (which translates into a logarithmic error of \variable{output/ape_NN_log.txt} dex)  which is far below typical observational errors and even smaller away from the extremes of parameter space (see Fig.~\ref{fig:ape_NN}). In fact, we will add additional observational uncertainty to our mock observational data later during training of the neural posteriro estimater network.
